---
title: "Vector Autoregression (VAR) Model"
author: "Ibikunle Gabriel"
date: '`r Sys.Date()`'
output:
  
  html_document: 
    theme:
      bootswatch: superhero
    toc: TRUE
  word_document: default
fig_height: 3
fig_width: 6 
fontsize: 9pt
---

## **What is a VAR Model?**
A Vector Autoregression (VAR) model is a time series model in which multiple variables depend on their own past values each other’s past values.
So instead of modeling one variable, you model a system of variables together.

**Loading required packages**

```{r setup, include=FALSE}
# Load the packages
library(vars)
library(tseries)
library(urca)
library(ggplot2)
library(reshape2)
library(dplyr)
library(tidyr)
library(bslib)
```


**Set and confirm the working directory**
```{r setup_root_dir, include=FALSE}
# Set working directory for all chunks during knitting
knitr::opts_knit$set(root.dir = "C:/Users/Hp/Documents/ARDL")
```

```{r}
getwd()
```


**load the data into the R environment **
```{r}
data <- read.csv("C:\\Users\\Hp\\Documents\\ARDL\\ECON DEVELOPMENT DATA.csv")
```


**Select the Variables needed and convert t time series object**
```{r}
data_var <- data[, c("ECNDEV","INF", "EDU", "K", "HEA", "UNEMP")]
```

```{r}
data_var <- ts(data_var)
```


**Determine Optimal Lag Length**
```{r}
lag_selection <- VARselect(data_var, lag.max = 10, type = "const")
lag_selection
```


```{r}
lag_selection$selection
```

**Estimate VAR Model**
```{r}
var_model <- VAR(data_var, p = 2, type = "const")

summary(var_model)
```

**Check Stability of VAR**
```{r}
stability <- roots(var_model)
stability

plot(stability)
```

*Diagnostic Tests*
**Serial Correlation**
```{r}
serial.test(var_model, lags.pt = 16, type = "PT.asymptotic")
```

**Heteroskedasticity**
```{r}
arch.test(var_model, lags.multi = 5)
```

**Normality**
```{r}
normality.test(var_model)
```

**Impulse Response Function (IRF)**

```{r}
irf_result <- irf(var_model,
                  impulse = "ECNDEV",
                  response = "HEA",
                  n.ahead = 10,
                  boot = TRUE)

plot(irf_result)
```
```{r}
irf_result1 <- irf(var_model,
                  impulse = "ECNDEV",
                  response = "HEA",
                  n.ahead = 10,
                  boot = TRUE)

plot(irf_result1)
```


```{r}
irf_result2 <- irf(var_model,
                   impulse = "ECNDEV",
                   response = "INF",
                   n.ahead = 10,
                   boot = TRUE)

plot(irf_result2)
```


```{r}
irf_result3 <- irf(var_model,
                   impulse = "ECNDEV",
                   response = "EDU",
                   n.ahead = 10,
                   boot = TRUE)

plot(irf_result3)
```


```{r}
irf_result4 <- irf(var_model,
                   impulse = "ECNDEV",
                   response = "K",
                   n.ahead = 10,
                   boot = TRUE)

plot(irf_result4)
```


```{r}
irf_result5 <- irf(var_model,
                   impulse = "ECNDEV",
                   response = "UNEMP",
                   n.ahead = 10,
                   boot = TRUE)

plot(irf_result5)

```

**Get all responses automatically**

```{r}
irf_all <- irf(var_model, n.ahead = 10, boot = TRUE)
plot(irf_all)
```

**Forecast Error Variance Decomposition (FEVD)**
```{r}
fevd_result <- fevd(var_model, n.ahead = 10)

fevd_result
# Plot FEVD
plot(fevd_result)
```


**Forecast (Optional)**
```{r, echo=FALSE, warning=FALSE, message=FALSE}
forecast_var <- predict(var_model, n.ahead = 10)
plot(forecast_var)
```

*------------ VAR with retail dataset-----------------*


```{r}
df1 <- read.csv("C:\\Users\\Hp\\Documents\\ARDL\\retail_sales_dataset.csv")
```


```{r}
# Select the Variables
data_var <- df1[, c("Age", "Price", "TotalAmount","Quantity")]
```


```{r}
# Convert to time series object
data_var <- ts(data_var)
```

```{r}
# Determine Optimal Lag Length
lag_selection <- VARselect(data_var, lag.max = 10, type = "const")
lag_selection
lag_selection$selection
```


```{r}
# Estimate VAR Model
var_model <- VAR(data_var, p = 2, type = "const")
summary(var_model)
```


```{r}
# Check Stability of VAR
stability <- roots(var_model)
stability
roots_var <- roots(var_model)
```

```{r}
# Plot empty coordinate system
plot(Re(roots_var), Im(roots_var),
     xlim = c(-1.5, 1.5),
     ylim = c(-1.5, 1.5),
     xlab = "Real Part",
     ylab = "Imaginary Part",
     main = "Stability of VAR Model",
     pch = 19,
     col = "blue")
# Add unit circle
theta <- seq(0, 2*pi, length = 500)
lines(cos(theta), sin(theta), col = "red", lwd = 2)
# Add horizontal and vertical axis
abline(h = 0, v = 0, col = "gray")

# Add grid
grid(col = "lightgray")

# Save the plot automatically to your working directory
ggsave("Stability_plot.png", width = 12, height = 8, dpi = 300)

```

**Diagnostic Tests**
```{r}
## Serial Correlation
serial.test(var_model, lags.pt = 16, type = "PT.asymptotic")
```


```{r}
## Heteroskedasticity
arch.test(var_model, lags.multi = 5)
```



```{r}
# Normality
normality.test(var_model)
```

**Impulse Response Function (IRF)**

```{r}

# Plot of just two variables
irf_result <- irf(var_model,
                  impulse = "Price",
                  response = "TotalAmount",
                  n.ahead = 10,
                  boot = TRUE)

plot(irf_result)
```
```{r}

# Plot of just two variables
irf_result1 <- irf(var_model,
                  impulse = "Price",
                  response = "TotalAmount",
                  n.ahead = 10,
                  boot = TRUE)

plot(irf_result1)
```
```{r}

# Plot of just two variables
irf_result2 <- irf(var_model,
                  impulse = "Price",
                  response = "Age",
                  n.ahead = 10,
                  boot = TRUE)

plot(irf_result2)
```

```{r}

# Plot of just two variables
irf_result3 <- irf(var_model,
                  impulse = "Price",
                  response = "Quantity",
                  n.ahead = 10,
                  boot = TRUE)

plot(irf_result3)
```

```{r}
# all responses automatically
irf_all <- irf(var_model, n.ahead = 10, boot = TRUE)
plot(irf_all)
```


```{r}
# all responses automatically
irf_all <- irf(var_model, n.ahead = 10, boot = TRUE)
plot(irf_all)
```

```{r}
# A more publishable plot
irf_all <- irf(var_model,
               n.ahead = 10,
               boot = TRUE,
               ci = 0.95)

irf_df <- do.call(rbind, lapply(names(irf_all$irf), function(imp) {
  
  responses <- irf_all$irf[[imp]]
  lower <- irf_all$Lower[[imp]]
  upper <- irf_all$Upper[[imp]]
  
  df <- as.data.frame(responses)
  df$Horizon <- 0:(nrow(df)-1)
  df$Impulse <- imp
  
  df_long <- pivot_longer(df,
                          cols = -c(Horizon, Impulse),
                          names_to = "Response",
                          values_to = "IRF")
  
  df_long$Lower <- as.vector(lower)
  df_long$Upper <- as.vector(upper)
  
  df_long
}))


ggplot(irf_df, aes(x = Horizon, y = IRF)) +
  geom_ribbon(aes(ymin = Lower, ymax = Upper), fill = "skyblue", alpha = 0.3) +
  geom_line(color = "darkblue", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_grid(Response ~ Impulse) +
  labs(title = "Impulse Response Functions",
       x = "Forecast Horizon",
       y = "Response") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# Save the plot automatically to your working directory
ggsave("IRF_plot.png", width = 12, height = 8, dpi = 300)
```


```{r}
par(mar = c(2, 2, 2, 2))  # default margins
# Forecast Error Variance Decomposition (FEVD)
fevd_result <- fevd(var_model, n.ahead = 10)

fevd_result
# Plot FEVD
plot(fevd_result)
```

```{r}
# A more publishable plot
fevd_df <- do.call(rbind, lapply(names(fevd_result), function(var) {
  
  df <- as.data.frame(fevd_result[[var]])
  df$Horizon <- 1:nrow(df)
  df$Variable <- var
  
  df
  
}))

fevd_long <- melt(fevd_df,
                  id.vars = c("Horizon", "Variable"),
                  variable.name = "Shock",
                  value.name = "Contribution")

ggplot(fevd_long, aes(x = Horizon, y = Contribution, fill = Shock)) +
  geom_area(position = "stack", alpha = 0.9) +
  facet_wrap(~Variable, scales = "free_y") +
  labs(title = "Forecast Error Variance Decomposition",
       x = "Forecast Horizon",
       y = "Variance Contribution",
       fill = "Shock Variable") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )+
  scale_fill_brewer(palette = "Set2")

# Save the plot automatically to your working directory
ggsave("FEVD_plot.png", width = 10, height = 6, dpi = 300)
```

```{r}
# Forecast (Optional)
forecast_var <- predict(var_model, n.ahead = 2)
par(mar = c(2, 2, 2, 2))
plot(forecast_var)
```

```{r}

# Forecast variance plot with ggplot
forecast_var <- predict(var_model, n.ahead = 10, ci = 0.95)
library(dplyr)
library(tidyr)

forecast_df <- do.call(rbind, lapply(names(forecast_var$fcst), function(var){
  
  df <- as.data.frame(forecast_var$fcst[[var]])
  
  df$Horizon <- 1:nrow(df)
  df$Variable <- var
  
  df
}))

library(ggplot2)

ggplot(forecast_df, aes(x = Horizon, y = fcst)) +
  
  geom_ribbon(
    aes(ymin = lower, ymax = upper),
    fill = "grey70",
    alpha = 0.4
  ) +
  
  geom_line(
    color = "green",
    linewidth = 1
  ) +
  
  facet_wrap(~Variable, scales = "free_y") +
  
  labs(
    title = "VAR Forecasts with 95% Confidence Intervals",
    x = "Forecast Horizon",
    y = "Forecast Value"
  ) +
  
  theme_classic(base_size = 14) +
  
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    strip.text = element_text(face = "bold")
  )
```


```{r}
forecast_var <- predict(var_model, n.ahead = 10, ci = 0.95)

forecast_df <- do.call(rbind, lapply(names(forecast_var$fcst), function(var){
  
  df <- as.data.frame(forecast_var$fcst[[var]])
  
  df$Horizon <- 1:nrow(df)
  df$Variable <- var
  
  df
}))

hist_df <- data.frame(data_var)

hist_df$Time <- 1:nrow(hist_df)

hist_long <- tidyr::pivot_longer(
  hist_df,
  cols = -Time,
  names_to = "Variable",
  values_to = "Value"
)


hist_df <- data.frame(data_var)

hist_df$Time <- 1:nrow(hist_df)

hist_long <- tidyr::pivot_longer(
  hist_df,
  cols = -Time,
  names_to = "Variable",
  values_to = "Value"
)

forecast_df$Time <- max(hist_long$Time) + forecast_df$Horizon

ggplot() +
  
  # historical data
  geom_line(
    data = hist_long,
    aes(x = Time, y = Value),
    color = "black",
    linewidth = 0.8
  ) +
  
  # confidence interval
  geom_ribbon(
    data = forecast_df,
    aes(x = Time, ymin = lower, ymax = upper),
    fill = "grey70",
    alpha = 0.4
  ) +
  
  # forecast line
  geom_line(
    data = forecast_df,
    aes(x = Time, y = fcst),
    color = "blue",
    linewidth = 1,
    linetype = "dashed"
  ) +
  
  facet_wrap(~Variable, scales = "free_y") +
  
  labs(
    title = "VAR Forecast with Historical Data",
    x = "Time",
    y = "Value"
  ) +
  
  theme_classic(base_size = 14) +
  
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    strip.text = element_text(face = "bold")
  )
```

## References

Pesaran, M. H., Shin, Y., & Smith, R. J. (2001).
Bounds testing approaches to the analysis of level relationships. Journal of Applied Econometrics, 16(3), 289–326.https://doi.org/10.1002/jae.616

Pesaran, M. H., & Smith, R. (1995).
Estimating long-run relationships from dynamic heterogeneous panels. Journal of Econometrics, 68(1), 79–113.https://doi.org/10.1016/0304-4076(94)01644-F
